### 2016-02-03 text/Chapter06.tex
Reorg chapter 4 and 5, adding unique typing to chapter 4

%!TEX root = ../main.tex

\chapter{Unified coeffect language ???} 
\label{ch:unified}

%===================================================================================================

\section{Introduction}

The main goal of this thesis is to provide a \emph{unified} calculus for tracking context 
dependence. We have not achieved this goal yet. In Chapter~\ref{ch:applications}, we identified 
two kinds of contextual properties that we further covered separately -- flat coeffects in 
Chapter~\ref{ch:flat} track whole-context properties and structural coeffects, covered in 
Chapter~\ref{ch:structural}, track per-variable properties. In this chapter, we unify
the two notions. We introduce a \emph{unified coeffect} calculus that generalizes the two systems
and can be instantiated to track both flat and structural properties (Section~\ref{sec:unified-unified}).

Although the results presented in this thesis are mainly of a theoretical nature, we indeed 
believe that coeffects should be integrated in main-stream programming languages. In the second
part of this chapter, we discuss an alternative approach to defining coeffect 
systems which highlights the relationship between our work and related work arising from modal logics 
(Section~\ref{sec:unified-meta}). Finally, we outline one possible approach for practical
implementations of coeffects (Section~\ref{sec:unified-impl}).

% ==================================================================================================

\section{The unified coeffect calculus}
\label{sec:unified-unified}

The flat coeffect calculus (Figure~\ref{fig:flat-types}) and the structural coeffect calculus 
differ in a number of ways (Figure~\ref{fig:struct-types}). Understanding the differences is the 
key to reconciling the two systems:

\begin{itemize}
\item Structural coeffect calculus contains explicit rules for context manipulation
  (weakening, contraction, exchange). In the flat coeffect calculus, these rules are not defined
  explicitly, but are admissible.

\item In the structural coeffect calculus, the variable context is treated as a vector
  and is annotated with a vector of (scalar) coeffects. In the flat coeffect calculus,
  the variable context is a set and is annotated with a single (scalar) coeffect.

\item In the flat coeffect calculus, we distinguish between splitting of the context requirements
  and merging of context requirements ($\cpar$ and $\czip$, respectively). In the structural
  coeffect calculus, the operations (which model splitting and appending vectors) are invertible 
  and so the structural coeffect algebra requires just $\atimes$.
\end{itemize}

\noindent
In the unified calculus presented in this section, we address the three differences as follows.
We use calculus with explicit rules for context manipulation. In systems that arise from flat
calculi, the rules can be applied freely without changing the coeffects. We generalize the 
structure of coeffect annotations using the notion of a ``container'' which can be specialized 
to obtain a single annotation or a vector of annotations. Finally, we distinguish between 
splitting and merging of context requirements (using the notation $\apar$ and $\azip$, 
respectively). For structural coeffect calculi, the two operators coincide, but for flat
coeffect calculi, they provide the needed flexibility.

\subsection{Shapes and containers}

Our notion of a \emph{coeffect container} is based on the idea of a container introduced by 
Abbott et al. \cite{types-containers}. Interestingly, the work on containers has later been linked
to comonads by Ahman et al. \cite{comonads-containers}. Intuitively a container describes data
types such as lists, trees or streams. A container is formed by shapes (\eg~lengths of lists).
For every shape, we can obtain a set of positions in the container (\eg~offsets in a list of a
specified length). More formally:

\begin{definition}
A \emph{container} $S \triangleleft P$ is given by a set $S$ of shapes and a shape-indexed family
$P : S \rightarrow \ident{Set}$ of positions.
\end{definition}

\noindent
Well-known examples of containers include lists, non-empty lists, (un\-boun\-ded) streams and 
singleton data type (which contains exactly one element). Two containers relevant to our work
are lists and singleton data types:

\begin{itemize}
\item The container representing lists is given by $S \triangleleft P$ where 
  shapes are integers $S = \ident{Nat}$ (lengths of a list). The set of positions for a 
  given length $n$ is a set of indices $P n = \{ 1 \ldots n \}$.

\item The contianer representing singleton data type is given by $S \triangleleft P$ where
  shapes are given by a singleton set $S = \{ \ast \}$ and the set of positions for the
  shape $\ast$ contains exactly one position $P \ast = \{ 0 \}$.
\end{itemize}

\noindent
In the unified coeffect calculus, the structure of coeffect annotations is defined by a 
container with additional operations (discussed later) that links it with the free-variable 
context $\Gamma$.

\subsection{Structure of coeffects}
In the structural coeffect calculus, the structure annotation was formed by a vector of coeffect
sclars. The unified coeffect calculus is similar, but a \emph{vector} is replaced with a 
\emph{container}. The primitive coeffect annotations in the unified calculus are formed by
a \emph{coeffect scalar}, which remains the same as in structural coeffect calculus
(Definition~\ref{def:structural-scalar}). In this section, we refer to it as
\emph{\cclrd{unified coeffect scalar}} (and we repeat the definition below). Then we
define \emph{\sclrd{unified coeffect containers}} which determines how coeffect scalar values 
are attached to the free-variable context. Finally, we define the \emph{\aclrd{unified coeffect 
algebra}} which consists of shape-indexed coeffect scalar values.

As in the structural coeffect calculus, the contexts in the unified calculus are annotated with 
shape-indexed coeffects, written as $\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash e : \tau$; 
functions take just a single input parameter and so are annotated with scalar coeffect values 
$\sigma \xrightarrow{\cclrd{r}} \tau$.

\paragraph{Coeffect scalar.}
The following definition of the coeffect scalar structure repeats the Definition~\ref{def:structural-scalar}
from the previous chapter.

\begin{definition}
A \emph{\cclrd{unified coeffect scalar}} $(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ is a set 
$\C$ together with elements $\cunit, \czero \in \C$, relation $\cleq$ and binary operations 
$\cseq, \cpar$ such that $(\C, \cseq, \cunit)$ and $(\C, \cpar, \czero)$ are monoids and
$(\C, \cleq)$ is a pre-order. That is, for all $r,s,t\in \C$:
%
\begin{equation*}
\begin{array}{ccr}
r \;\cseq\; (s \;\cseq\; t) = (r \;\cseq\; s) \;\cseq\; t  &
\cunit \;\cseq\; r = r = r \;\cseq\; \cunit &
\textnormal{(monoid)}   
\\
r \;\cpar\; (s \;\cpar\; t) = (r \;\cpar\; s) \;\cpar\; t &
\czero \;\cpar\; r = r = r \;\cpar\; \czero &
\textnormal{(monoid)}   
\\
\textnormal{if}~~r\; \cleq\; s ~~\textnormal{and}~~s\; \cleq\; t~~\textnormal{then}~~r\; \cleq\; t&
t\; \cleq\; t &
\textnormal{(pre-order)}   
\end{array}
\end{equation*}
\end{definition}

\noindent
As previously, the monoid $(\C,\cseq,\cunit)$ models sequential composition; the laws guarantee
an underlying category structre; $\cunit$ and $\czero$ represent an accessed and unused variable, 
respectively.

The $\cpar$ operation models combining of context requirements arising from multiple parts of a 
program. The meaning depends on the coeffect container. The operation can either combine requirements
of individual variables (structural) or requirements attached to the whole context of multiple
sub-exp\-ressions (flat).

\paragraph{Coeffect containers.}
The coeffect container is a container that determines how are scalar coeffect annotations attached 
to free-variable contexts. In addition to a container $\SHP$ with shapes and shape-indexed positions, the
coeffect container also provides a mapping that returns the shape of a free-variable context. The
mapping between the shape of the variable context and the shape of the coeffect annotation
is not necessarily bijective. For example, coeffect annotations in flat systems have just a single 
shape $\SH = \{ \ast \}$.

In the coeffect judgment $\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash e : \tau$, the coeffect annotation
$\aclrd{\textbf{r}}$ is drawn from the set of coeffect scalars $\C$ indexed by the shape of $\Gamma$.
We write $\sclrd{s} = \slen{\Gamma}$ for the shape corresponding to $\Gamma$. The operation $\SP\sclrd{s}$
returns a \emph{set} of positions and so we can write $\aclrd{\textbf{r}} \in \SP\sclrd{s} \rightarrow \C$ 
as a mapping from positions (defined by the shape) to scalar coeffects. We usually write this as 
the exponent $\aclrd{\textbf{r}} \in \C^{\SP\sclrd{s}}$.

The coeffect container is also equipped with an operation that appends shapes (when we concatenate
variable contexts) and two special shapes in $\SH$ representing empty context and singleton context.

\begin{definition}
A \emph{\sclrd{coeffect container}} structure $(\SHP, \stimes, \sempty, \sunit, \slen{-})$ 
comprises a container $\SHP$ with a binary operation $\stimes$ on $\SH$ for appending shapes, a 
mapping from free-variable contexts to shapes $\slen{\Gamma} \in \SH$, and elements $\sempty,\sunit \in
\SH$ such that $(\SH, \stimes, \sempty)$ is a monoid.

The elements $\sempty$ and $\sunit$ represent the shapes of empty and singleton free-variable contexts 
respectively. The $\stimes$ operation corresponds to concatentation of free-variable contexts. Given
$\Gamma_1$ and $\Gamma_2$ such that $\sclrd{s_1}=\slen{\Gamma_1}, \sclrd{s_2}=\slen{\Gamma_2}$, 
we require that $\sclrd{s_1}\,\stimes\,\sclrd{s_2}=\slen{\Gamma_1, \Gamma_2}$. 
\end{definition}

\noindent
As said earlier, we use two kinds of coeffect containers that describe the structure of vectors 
(for structural coeffects) and the shape of trivial singleton container (for flat coeffects):

\begin{example}
\label{ex:unified-struct-shape}
Structural coeffect container is defined as $\sclrd{(\SHP, {|}\textnormal{--}{|}, +, 0, 1)}$
where $\SH=\mathbb{N}$ and $\SP \sclrd{n}=\{ 1 \ldots n \}$. The shape mapping ${|}\Gamma{|}$ returns the 
number of variables in $\Gamma$. Empty and singleton contexts are annotated with $0$ and $1$, 
respectively, and shapes of combined contexts are added so that ${|}\Gamma_1, \Gamma_2{|} = 
{|}\Gamma_1{|} + {|}\Gamma_2{|}$.

Therefore, a coeffect annotation is a \emph{vector} 
$\aclrd{\textbf{r}} \in \C^{\SP\sclrd{n}}$ and assigns a coeffect scalar $\aclrd{\textbf{r}}(i) \in \C$ 
for each position (corresponding to a variable $x_i$ in the context).
\end{example}

\begin{example}
\label{ex:unified-flat-shape}
Flat coeffect container is defined as $\sclrd{(\SHP, {|}-{|}, \diamond, \sflat, \sflat)}$.
The container is defined as a singleton data type $\SH = \{\sflat \}$ and $\SP \sflat = \{ 0 \}$
with a constant function ${|}\Gamma{|}=\sflat$ and a trivial operation $\sflat\,\diamond\,\sflat = \sflat$.

That is, there is a single shape
$\sflat$ with a single position and all free-variable contexts have the same singleton shape.
Therefore, a coeffect annotation is drawn from $\C^\{\sflat\}$ which is isomorphic to $\C$
and so a coeffect scalar $\cclrd{r}\in \C$ is associated with every free-variable context.
\end{example}

\newpage
\begin{example}
Similarly to the previous example, we can also define a coeffect container with \emph{no} positions,
\ie~$\sclrd{(\SHP, {|}-{|}, \diamond, \sflat, \sflat)}$ where $\SH = \{\sflat \}$, $\SP \sflat = \emptyset$,
${|}\Gamma{|}=\sflat$ and $\sflat\,\diamond\,\sflat = \sflat$.

This reduces our system to the simply-typed $\lambda$-calculus with no context annotations, because
$\SP\sflat=\emptyset$ and so coeffect annotations would be from the set $\C^\emptyset$.
\end{example}

\paragraph{Unified coeffect algebra.}
The coeffect calculus annotates judgments with shape-indexed coeffect annotations.
The \emph{\aclrd{unified coeffect algebra}} combines a coeffect scalar and coeffect container
to define shape-indexed coeffects and operations for manipulating these.

The definition here reconciles the third point discussed in Section~\ref{sec:unified-unified} --
the fact that flat coeffects use separate operations for splitting and merging ($\cpar$ and $\czip$)
while structural coeffects use tensor $\atimes$. In the unified calculus, we use two operators
that can, however, be coincide.

\begin{definition}
Given a \cclrd{unified coeffect scalar} $(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ and a 
\sclrd{coeffect container} $(\SHP, \slen{-}, \stimes, \sempty, \sunit)$ a \emph{\aclrd{unified 
coeffect algebra}} extends the two structures with $(\azip, \apar, \azero)$ where $\azero \in \C^{\SP \sempty}$
is a coeffect annotation for the empty context and $\azip, \apar$ are families of operations that 
combine coeffect annotations indexed by shapes. That is $\forall \sclrd{n},\sclrd{m} \in \SH$:
%
\begin{equation*}
\apar_{\sclrd{m},\sclrd{n}}, \azip_{\sclrd{m},\sclrd{n}} ~:~ \C^{\SP \sclrd{m}} \times \C^{\SP \sclrd{n}} \rightarrow \C^{\SP (\sclrd{m}\,\stimes\,\sclrd{n})}
\end{equation*}
\end{definition}

\noindent
A coeffect algebra induces the following three additional operations:
%
\begin{equation*}
\begin{array}{rclcrcl}
 \alift{-} &\narrow{:}& \C \rightarrow \C^{\SP \sunit}  \\
 \alift{\cclrd{x}} &\narrow{=}& \lambda \_ . \cclrd{x}   \\
\\[-0.5em]
 \aseq_{\sclrd{m}} &\narrow{:}&  \C \times \C^{\SP \sclrd{m}} \rightarrow \C^{\SP \sclrd{m}}   \\
 \cclrd{r}\, \aseq\, \aclrd{\textbf{s}} &\narrow{=}& \lambda \sclrd{s} . \cclrd{r}\, \cseq\, \aclrd{\textbf{s}}(\sclrd{s}) \\
\\[-0.5em]
 \alen{-} &\narrow{:}& \C^{\SP \sclrd{m}} \rightarrow \sclrd{m}\\
 \alen{\aclrd{\textbf{r}}} &\narrow{=}& \sclrd{m}
\end{array}
\end{equation*}
%
The $\alift{-}$ operation lifts a scalar coeffect to a shape-indexed coeffect that is indexed by 
the singleton context shape. The $\aseq_{\sclrd{m}}$ operation is a left multiplication of a 
vector by scalar. As we always use bold font for vectors and ordinary font for scalars (as well 
as distinct colour), using the same symbol is not ambiguous. We also tend to omit the subscript
$\sclrd{m}$ and write $\aseq$.

Finally, we define $\alen{}$ as an operation that returns the shape of a given shape-indexed 
coeffect. The only purpose is to simplify notation, as we often need to specify that shapes of 
variable context and coeffect match, \eg~$\alen{\aclrd{\textbf{r}}} = \slen{\Gamma}$.

\paragraph{Splitting and merging coeffects.}
The operators $\apar$ and $\azip$ combine shape-indexed coeffects associated with two contexts. 
For example, assume we have $\Gamma_1$ and $\Gamma_2$ with coeffects $\aclrd{\textbf{r}} \in 
\C^{\SP\sclrd{m}}$ and $\aclrd{\textbf{s}} \in \C^{\SP\sclrd{n}}$. In the structural system, the 
context shapes $\sclrd{m}, \sclrd{n}$ denote the number of variables in the two contexts. The 
combined context $\Gamma_1, \Gamma_2$ has a shape $\sclrd{m}\,\stimes\,\sclrd{n}$ and the combined 
coeffects $\aclrd{\textbf{r}} \,\azip\, \aclrd{\textbf{s}}, \aclrd{\textbf{r}} \,\apar\, \aclrd{\textbf{s}} 
\in \C^{\SP(\sclrd{m} \stimes \sclrd{n})}$ are indexed by that shape.

For structural coeffect systems such as bounded reuse, both $\apar$ and $\azip$ are
just the tensor product $\times$ of vectors. For flat coeffect systems, the operations
can be defined independently, letting $\azip=\czip$ and $\apar=\cpar$.

The difference between $\azip$ and $\apar$ is clarified by the semantics (Sec\-tion~\ref{sec:unified-semantics}), 
where $\aclrd{\textbf{r}}\,\azip\,\aclrd{\textbf{s}}$ is an annotation of the \emph{codomain} of 
a morphism that merges the capabilities provided by two contexts (in the syntactic 
reading, splits the context requirements), while $\aclrd{\textbf{r}}\,\apar\,\aclrd{\textbf{s}}$ 
is an annotation of the \emph{domain} of a morphism that splits the capabilities of a single 
context into two parts (in the syntactic reading, merges their context requirements). 
Syntactically, this means that we always use $\azip$ in the rule \emph{assumptions} and 
$\apar$ in \emph{conclusions}.

>>>>

\chapter{The structural coeffect calculus} 
\label{ch:structural}

In this chapter, we take the next step towards a unified theory of context-aware programming
languages. As already discussed, we attempt to find abstractions that are \emph{general enough} 
to capture a wide range of useful programming language features, but \emph{specific enough} to 
let us identify and capture interesting properties of the context.

In Chapter~\ref{ch:applications}, we discussed two notions of context. Context-aware programming
languages that that capture whole-context properties were generalized by the \emph{flat coeffect 
calculus} in Chapter~\ref{ch:flat}. Here, we consider per-variable contextual properties
and we introduce the \emph{structural coeffect calculus}.

The flat coeffect system captures a number of interesting use-cases (implicit parameters, liveness 
and data-flow), but provides relatively imprecise approximation and weak syntactical properties. 
Some of those examples can be also seen as per-variable properties. For those, structural coeffect
systems give a more precise information about the context. However, we also look at other applications 
that arise from the work on sub-structural logics, a pathway to coeffects discussed in 
Section~\ref{sec:path-logic}.

As we will see, the categorical semantics of the structural coeffect calculus is based on the
same underlying structure as the semantics of the flat calculus. However, the syntactical properties
of the calculus are quite different and so it is worth considering it in a separate chapter.

\paragraph{Chapter structure and contributions}
\begin{itemize}
\item We present a \emph{structural coeffect calculus} as a type system that is parameterized 
  by a \emph{structural coeffect algebra} (Section~\ref{sec:struct-calculus}. We show how the 
  system captures fine-grained liveness and data-flow information, as well as a calculus for bounded 
  reuse (checking how many times is a variable accessed).

\item We extend indexed comonads introduced in the previous chapter to 
  \emph{structural indexed comonads} and use them to provide the semantics of structural 
  coeffect calculus (Section~\ref{sec:struct-semantics}). As with the flat version, the
  semantics provides a motivation for the design of the calculus.

\item We discuss the equational theory of the calculus. For structural coeffects, the substitution 
  lemma holds for all instances of the calculus. Thus, we prove the type-preservation property 
  for all structural calculi for both call-by-name and call-by-value (Section~\ref{sec:struct-syntax}). 
\end{itemize}

% ==================================================================================================
%                                                                                      
%     ###           #                       #                 #       #                 
%      #            #                       #                 #                         
%      #    # ##   ####   # ##    ###    ## #  #   #   ###   ####    ##     ###   # ##  
%      #    ##  #   #     ##  #  #   #  #  ##  #   #  #   #   #       #    #   #  ##  # 
%      #    #   #   #     #      #   #  #   #  #   #  #       #       #    #   #  #   # 
%      #    #   #   #  #  #      #   #  #  ##  #  ##  #   #   #  #    #    #   #  #   # 
%     ###   #   #    ##   #       ###    ## #   ## #   ###     ##    ###    ###   #   # 
%                                                                                     
% ==================================================================================================

\section{Introduction}
\label{sec:structural-intro}

Compared to the previous chapter, structural coeffect calculi we consider are more 
homogeneous and so finding the common pattern is in some ways easier. However, the systems
are more complicated as they need to keep annotations attached to individual variables
and thus require explicit structural rules. Before looking at the system, we briefly consider
the most important related work.

%---------------------------------------------------------------------------------------------------

\subsection{Related work}

In the previous chapter, we discussed the correspondence between coeffects and effects (and 
between comonads and monads). As noted earlier, the $\lambda$-calculus is asymmetric in that
an expression has multiple inputs (variables in the context), but just a single result 
(the resulting value) and so monads and effects have no notion directly corresponding to 
structural coeffect systems.

The work in this chapter is more closely related to sub-structural type systems \cite{substruct-attpl-intro}.
While sub-structural systems remove some or all of \emph{weakening}, \emph{contraction} and
\emph{exchange} rules, our systems keep them, but use them to manipulate both the context
and its annotations.

Our work follows the ``language semantics'' style in that we provide a structural semantics to
the terms of ordinary $\lambda$-calculus. By contrast the most closely related work has been done in 
the meta-language style, which extends the terms and types of a language with constructs for explicitly
manipulating the context. This includes Contextual Modal Type Theory (CMTT) \cite{logic-cmtt}, where variables
may be of a type $A[\Psi]$ denoting a value of type $A$ that requires context $\Psi$. In CMTT, 
$A[\Psi]$ is a first-class type, while structural coeffect systems do not expose coeffect 
annotations as stand-alone types (indexed comonads only appear in the semantics).

Our structural coeffect systems annotate the whole variable context with a \emph{vector} of annotations.
For example, a context with variables $x$ and $y$ annotated with $\cclrd{s}$ and $\cclrd{t}$, 
respectively is written as $\coctx{x\!:\!\tau_1,y\!:\!\tau_2}{\alift{\cclrd{s},\cclrd{t}}}$.
A benefit of this approach is that the typing judgements have the same structure as those of the 
flat coeffect calculus. As discussed in Chapter~\ref{ch:unified}, this makes it possible to unify 
the two systems.

% =================================================================================================
%                                                            
%       ###           ##                   ##                 
%      #   #           #                    #                 
%      #       ###     #     ###   #   #    #    #   #   ###  
%      #          #    #    #   #  #   #    #    #   #  #     
%      #       ####    #    #      #   #    #    #   #   ###  
%      #   #  #   #    #    #   #  #  ##    #    #  ##      # 
%       ###    ####   ###    ###    ## #   ###    ## #  ####  
%                                                           
% =================================================================================================

\section{Structural coeffect calculus}
\label{sec:struct-calculus}

In the structural coeffect calculus, a vector of variables forming the free-variable context is 
annotated with a vector of primitive (scalar) coeffect annotations. These annotations differ
for different coeffect calculi and their properties are captured by the definition of \emph{structural 
coeffect scalar} below. The scalar annotations can be \eg~integers (how many past values we 
need) or values of a two-point lattice specifying whether a variable is live or not.

Scalar annotations are written as $\cclrd{r}, \cclrd{s}, \cclrd{t}$ (following the style used in 
the previous chapter). Functions always have exactly one input variable and so they are annotated
with a single coeffect scalar. Thus the expressions and types of structural coeffect calculi are the 
same as in the previous chapter. The only difference is that $\cclrd{r}, \cclrd{s}, \cclrd{t}$ now
range over values of a \emph{structural coeffect scalar} rather than over values of a \emph{flat 
coeffect algebra} (the definitions are related, but not the same).
%
\begin{equation*}
\begin{array}{rcl}
e &::=& x \sep \lambda x.e \sep e_1~e_2 \sep \kvd{let}~x = e_1~\kvd{in}~e_2\\
\tau &::=& T \sep \tau_1 \xrightarrow{\cclrd{r}} \tau_2
\end{array}
\end{equation*}
%
In the previous chapter, the free variable context $\Gamma$ was treated as a set. In the 
structural coeffect calculus, the order of variables matters. Thus we treat a free variable 
context as a vector with a uniqueness condition. We also write $\xlen{-}$ for the length
of the vector:
%
\begin{equation*}
\begin{array}{rcl}
 \Gamma &\narrow{=}& \langle x_1\!:\!\tau_1, \ldots, x_n\!:\!\tau_n \rangle \qquad \textnormal{such that}~~\forall i, j ~.~ i \neq j \implies x_i \neq x_j\\
 \xlen{\Gamma} &\narrow{=}& n
\end{array}
\end{equation*}
%
For readability, we use the usual notation $x_1\!:\!\tau_1, \ldots, x_1\!:\!\tau_1 \vdash e : \tau$ for
typing judgements, but the free variable context should be understood as a vector.
Furthermore, the notation $\Gamma_1, \Gamma_2$ is here seen as the tensor product.
Given $\Gamma_1 = \langle x_1\!:\!\tau_1, \ldots, x_n\!:\!\tau_n \rangle $ and
$\Gamma_2 = \langle x_{n+1}\!:\!\tau_{n+1}, \ldots, x_m\!:\!\tau_m \rangle $ then
$\Gamma_1, \Gamma_2 = \Gamma_1 \times \Gamma_2 = \langle x_1\!:\!\tau_1, \ldots, x_m\!:\!\tau_m \rangle$.

Free variable contexts are annotated with vectors of structural coeffect scalars. In what follows,
we write the vectors of coeffects as $\alift{\cclrd{r_1}, \ldots, \cclrd{r_n}}$. Meta-variables ranging 
over vectors are written as $\aclrd{\textbf{r}}, \aclrd{\textbf{s}}, \aclrd{\textbf{t}}$ (using bold face 
and colour to distinguish them from scalar meta-variables) and the length of a coeffect vector
is written as $\alen{\aclrd{\textbf{r}}}$. The structure for working with vectors
of coeffects is provided by the definition of \emph{structural coeffect algebra} below.

\subsection{Structural coeffect algebra}

The \emph{structural coeffect scalar} structure is similar to \emph{flat coeffect algebra} with the 
exception that it drops the $\czip$ operation. It only provides a monoid $(\C, \cseq, \cunit)$
modelling sequential composition of computations and a monoid $(\C, \cpar, \czero)$ representing
pointwise composition, as well as a relation $\cleq$ that defines sub-coeffecting.

\begin{definition}
\label{def:structural-scalar}
A \emph{\cclrd{structural coeffect scalar}} $(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ is a set 
$\C$ together with elements $\cunit, \czero \in \C$, relation $\cleq$ and binary operations 
$\cseq, \cpar$ such that $(\C, \cseq, \cunit)$ and $(\C, \cpar, \czero)$ are monoids and
$(\C, \cleq)$ is a pre-order. That is, for all $r,s,t\in \C$:
%
\begin{equation*}
\begin{array}{ccr}
r \;\cseq\; (s \;\cseq\; t) = (r \;\cseq\; s) \;\cseq\; t  &
\cunit \;\cseq\; r = r = r \;\cseq\; \cunit &
\textnormal{(monoid)}   
\\
r \;\cpar\; (s \;\cpar\; t) = (r \;\cpar\; s) \;\cpar\; t &
\czero \;\cpar\; r = r = r \;\cpar\; \czero &
\textnormal{(monoid)}   
\\
\textnormal{if}~~r\; \cleq\; s ~~\textnormal{and}~~s\; \cleq\; t~~\textnormal{then}~~r\; \cleq\; t&
t\; \cleq\; t &
\textnormal{(pre-order)}   
\\
\end{array}
\end{equation*}
%
The following distributivity axioms hold:
\begin{align*}
\quad (\cclrd{r}\, \cpar\, \cclrd{s}) \;\cseq\; \cclrd{t} & = (\cclrd{r} \,\cseq\, \cclrd{t}) \;\cpar\; (\cclrd{s}\, \cseq\, \cclrd{t}) \\
\quad \cclrd{t} \;\cseq\; (\cclrd{r}\, \cpar\, \cclrd{s}) & = (\cclrd{t} \,\cseq\, \cclrd{r}) \;\cpar\; (\cclrd{t}\, \cseq\, \cclrd{s})
\end{align*}
%
In addition, we require the following two properties of $\czero, \cpar$ and $\cleq$:
\begin{equation*}
\begin{array}{l}
 \quad \cclrd{r}\,\cleq\,\cclrd{r'} \; \Rightarrow \; \forall\cclrd{s}.\;(\cclrd{r}\,\cseq\,\cclrd{s})\;\cleq\;(\cclrd{r'}\,\cseq\,\cclrd{s}) \\
 \quad \czero\;\cleq\;(\czero\,\cseq\,\cclrd{r})
\end{array}
\end{equation*}
\end{definition}

\noindent
In the flat coeffect calculus, we used the $\czip$ operation to merge the annotations of contexts
available from the declaration site and the call site or, in the syntactic reading, to split the 
context requirements. The other difference is that the $\cpar$ operation of structural coeffect
scalar is not required to be idempotent. The bounded reuse example uses $\cpar$ which is not 
idempotent.

Finally, we now also require two properties that relate the ordering $\cleq$ with coeffects composed
via $\cpar$ and with $\czero$. These were previously required for bottom-pointed substitution lemma 
(Lemma~\ref{thm:cbn-substitution-bot}). Here, we require them for all structural coeffects as it holds
for all our examples; it is needed for the structural version of the substitution lemma 
(Lemma~\ref{thm:structural-substitution}).

In the structural coeffect calculus, we use a vector instead -- combining and splitting of 
coeffects becomes just vector concatenation or splitting, respectively; these are provided
by the tensor product. The operations on vectors are indexed by integers representing the lengths
of the vectors. The additional structure required by the type system for structural coeffect 
calculi is given by the following definition.

\begin{definition}
A \emph{\aclrd{structural coeffect algebra}} is formed by a structural coeffect scalar 
$(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ equipped with the following additional structures:

\begin{itemize}
\item Coeffect vectors $\aclrd{\textbf{r}}, \aclrd{\textbf{s}}, \aclrd{\textbf{t}}$, ranging over
  structural coeffect scalars indexed by vector lengths $\sclrd{m}, \sclrd{n} \in \sclrd{\mathbb{N}}$.

\item An operation that constructs a vector from scalars indexed by the vector length
  $\alift{-}_{\sclrd{n}} : \C \times \ldots \times \C \rightarrow \C^{\sclrd{n}}$
  and an operation that returns the vector length such that 
  $\alen{\aclrd{\textbf{r}}} = \sclrd{n}$ for $\aclrd{\textbf{r}} : \C^{\sclrd{n}}$

\item A pointwise extension of the $\cseq$ operator written as $\cclrd{t} \,\aseq\, \aclrd{\textbf{s}}$
  such that\\ $\cclrd{t} \,\aseq\, \alift{\cclrd{r_1}, \ldots, \cclrd{r_n}} = 
    \alift{\cclrd{t} \,\cseq\,\cclrd{r_1}, \ldots, \cclrd{t} \,\cseq\,\cclrd{r_n}}$.

\item An indexed tensor product $\atimes_{\sclrd{n},\sclrd{m}} : \C^{\sclrd{n}} \times \C^{\sclrd{m}} \rightarrow \C^{\sclrd{n}+\sclrd{m}}$
  that is used in both directions -- for vector concatenation and for splitting -- 
  which is defined as $\alift{\cclrd{r_1}, \ldots, \cclrd{r_n}} 
     \atimes_{\sclrd{n}, \sclrd{m}} \alift{\cclrd{s_1}, \ldots, \cclrd{s_m}}
   = \alift{\cclrd{r_1}, \ldots, \cclrd{r_n}, \cclrd{s_1}, \ldots, \cclrd{s_m}}$
\end{itemize}
\end{definition}

\noindent
The fact that the tensor product $\atimes_{\sclrd{n},\sclrd{m}}$ is indexed by the lengths
of the two vectors means that we can use it unambiguously for both concatenation of vectors
and for splitting of vectors, provided that the lengths of the resulting vectors are known.
In the following text, we usually omit the indices and write just 
$\aclrd{\textbf{r}} \atimes \aclrd{\textbf{s}}$, because the lengths of the coeffect vectors 
can be determined from the lengths of the matching free variable context vectors.

More generally, we could see the the coeffect annotations as \emph{containers} 
\cite{types-containers}. This approach is used in Chapter~\ref{ch:unified} to unify flat 
and structural systems.

\subsection{Structural coeffect types}
\label{sec:struct-calculus-types}

The type system for the structural coeffect calculus is similar to sub-structural type systems
in how it handles free variable contexts. The \emph{syntax-driven} rules do not implicitly allow
weakening, exchange or contraction -- this is done by checking the types of sub-expressions
in disjoint parts of the free variable context. Unlike in sub-structural logics, our system
does allow weakening, exchange and contraction, but via explicit \emph{structural} rules that
perform corresponding transformation on the coeffect annotation.

\paragraph{Syntax-driven rules.}
The syntax-driven rules of the type system are shown in Figure~\ref{fig:struct-types} (a).
The variable access rule (\emph{var}) annotates the corresponding variable as being accessed
using $\cunit$. Note that, as in sub-structural systems, the free variable context contains 
\emph{only} the accessed variable. Other variables can be introduced using explicit weakening.
Constants (\emph{const}) are type checked in an empty variable context, which is annotated with
an empty vector of coeffect annotations.

The (\emph{abs}) rule assumes that the free variable context of the body can be split into
a potentially empty \emph{declaration site} and a singleton context containing the bound
variable. The corresponding splitting is performed on the coeffect vector, uniquely associating
the annotation $\cclrd{s}$ with the bound variable $x$. This means that the typing rule removes
non-determinism of the type inference present in flat coeffect systems.

In (\emph{app}), the sub-expressions $e_1$ and $e_2$ use free variable contexts $\Gamma_1, \Gamma_2$
with coeffect vectors $\aclrd{\textbf{r}}, \aclrd{\textbf{s}}$, respectively. The function
value is annotated with a coeffect scalar $\cclrd{t}$.
The coeffect annotation of the composed expression is obtained by combining the annotations
associated with variables in $\Gamma_1$ and $\Gamma_2$. Variables in $\Gamma_1$ are only
used to obtain the function value, resulting in coeffects $\aclrd{\textbf{r}}$. The 
variables in $\Gamma_2$ are used to obtain the argument value, which is then sequentially
composed with the function, resulting in $\cclrd{t}\,\aseq\,\aclrd{\textbf{s}}$.

\paragraph{Structural rules.}
The remaining rules, shown in Figure~\ref{fig:struct-types} (b), are not syntax-directed.
They allow different transformation of the free variable context. We include (\emph{sub}) 
as one of the rules, allowing sub-coeffecting on coeffect scalars belonging to individual 
variables. The remaining rules capture \emph{weakening}, \emph{exchange} and \emph{contraction} 
known from sub-structural systems.

Rule (\emph{weak}) allows adding a variable to the context, extending the coeffect
vector with $\czero$ to mark it as unused, (\emph{exch}) provides a way to rearrange variables
in the context, performing the same reordering on the coeffect vector. Finally recall that 
variables in the free variable context are required to be \emph{unique}. The (\emph{contr}) rule 
allows re-using a variable. We can type check sub-expressions using two separate variables and 
then unify them using substitution. The resulting variable is annotated with $\cpar$ and it
is the only place in the structural coeffect system where context requirements are combined
(semantically, this is where the available context is shared).

%---------------------------------------------------------------------------------------------------
